Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-28 paper_nips/paper.tex
updated discussion structure

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.
%
Existing neuroimaging methods typically perform either discovery of unknown
neural structure or testing of neural structure associated with mental tasks.
%
However, modelling the neural correlates of mental tasks hinges
on the pertinence of the assumed neurobiological structure.
%XXX: this last statement is not cristal clear.
%
% XXX: Big gap between these two sentences. It is unclear what you call 'supervised' or 'unsupervised'.
%
We therefore propose to solve the unsupervised dimensionality reduction
and supervised task classification in
an identical statistical learning problem.
%
We show that this approach yields more accurate and more interpretable
neural models of psychological tasks in a reference dataset.
%

@@

\section{Introduction}
%
Methods used in neuroimaging research can be grouped by discovering
neurobiological structure or assessing the neural correlates associated
with mental tasks.
To discover spatial distributions of activation structure across time,
independent component analysis (ICA) \cite{beckmann2005} is often used
to decompose the BOLD (blood-oxygen level-dependent) signals into the
main modes of variation.
The ensuing spatial activation patterns are believed to represent
brain networks of
functionally interacting brain regions \cite{smith2009}.
Similarly, sparse principal component analysis (SPCA) \cite{varoqu2011}
has been used to
separate BOLD signals into parsimonious network components.
Thus extracted brain networks have been shown to be
manifestations of electrophysiological oscillation frequencies \cite{hipp15}.
Their fundamental organizational role is further
attested by continued covariation during sleep and anesthesia.
%
Network discovery is typically performed by applying ICA or SPCA on
unlabeled "resting-state" data. These capture brain dynamics
during ongoing random thought without controlled environmental stimulation.
A large proportion of the BOLD signal is known
not to correlate with a particular behavior, stimulus, or experimental task
\cite{fox07}.

On the other hand, to investigate
the neural correlates underlying mental tasks,
the general linear model (GLM) is the dominant approach \cite{friston94}.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI)
elucidate the functional interactions between voxels as a function
of experimental tasks \cite{friston97}.
Dynamic causal modeling (DCM), in turn, quantifies directed,
task-driven influences between regions
by treating the brain as a nonlinear dynamic system with unknown
neuronal states \cite{stephan04}.
As a last example, always more neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to labeled task data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses operate
without exception in a voxel space. 
%XXX:not exactly: the DCM is based on ROIs typically. PPI also ?

This ignores the fact that the BOLD
signal exhibits distributed spatial activation patterns.
Second, existing neuroimaging analyses do not acknowledge the fact that the
task-induced changes of the BOLD signal typically amount to less than 5\%
of baseline activity \cite{fox07}. 
%
% XXX I think that this is wrong. Moreover, it is not completely  clear how crucial this is. I think that there are better arguments to introduce the joint analysis of rest and task.
% For example, the existence of abundant RS data is an important practical reson to consider them. 'Better sample the manifold of brain states'
% 
They do thus not exploit the high similarity
of BOLD dynamics in the human brain at rest and during experimental tasks,
as observed using ICA \cite{smith2009}.

Both these biological properties can be conjointly exploited in a mixed
 (i.e., use rest and task data)
low-rank (i.e., perform network decomposition)
approach.
%
The integration of brain-network discovery into a 
supervised classification formally yields a semi-supervised learning framework that should identify the most relevant
neurobiological structure for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
Autoencoders
are one-layered learning models that condense the input data to
local and global representations
by improving reconstruction from them.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
%
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \caption {\textbf{Architecture. XXX develop the cpation. Main msg ?}
  }
\end{wrapfigure}
%
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and feeding the first-layer activation into any nonlinear convex
%% ICA contrasts are rarely convex ?!?
function \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Hence, autoencoders can flexibly project the neuroimaging data
onto the main axes of variation and thus
reverse-engineer properties of the underlying
neural processes \cite{olshausen96}.

In the present investigation,
a linear autoencoder will be fed by (unlabeled) rest data and
integrated as bottleneck
into a low-rank logistic regression fed by (labeled) task data.
% Using the chain rule for gradient descent, 
We can then solve the unsupervised data representation and the
supervised classification in an identical statistical learning
problem.
%
From the perspective of dictionary learning, the first layer represents
the discovered set of basis functions
whose linear combinations are learned
by the second layer \cite{olshausen96}.
%
Neurobiologically, this allows 
delineating a low-dimensional manifold of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
put probability mass on the most neurobiologically
valid models in the hypothesis set.
%
The generalization performance should consequently be improved due to 
reduced Vapnik-Chervonenkis dimensions of the trained estimator.
% I would not make direct claims on *VC* dimension, since you don't consider it explictly
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.

\section{Methods}
%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose the Human Connectome Project (HCP) resources
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500 unrelated,
healthy HCP participants.
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants.
The task paradigms include
1) working memory/cognitive control processing, 2)
incentive processing, 3) visual and somatosensory-motor processing,
4) language processing (semantic and phonological processing),
5) social cognition, 6) relational processing, and 7) emotional
processing. All data were acquired on the same Siemens Skyra 3T scanner.
Whole-brain EPI acquisitions were acquired with a
32 channel head coil (TR=720ms, TE=33.1ms, flip angle=52°, BW=2290Hz/Px,
in-plane FOV=$280\times180$mm, 72 slices, 2.0mm isotropic voxels).
The ``minimally preprocessed'' pipeline includes
gradient unwarping, motion correction, fieldmap-based EPI distortion
correction, brain-boundary-based registration of EPI to structural
T1-weighted scans, nonlinear (FNIRT) registration into MNI space,
and grand-mean intensity normalization. Activation maps were spatially
smoothed with a Gaussian kernel of 4mm (FWHM). A GLM was
implemented by FILM from the FSL suite with model regressors from convolution
with a “canonical” hemodynamic response function and from temporal derivatives.
HCP tasks were conceived to modulate activation
in a maximum of different brain regions and neural systems. Indeed, at
least 70\% of the participants showed consistent brain activity in
contrasts from the task battery, which certifies excellent
coverage \cite{barch2013}.
%
% No, this is not what is called 'coverage'
%
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common 60x72x60 space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supservised analyses were based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

@@

We were further interested in the utility of the optimal low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classifcation problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$19.2\times19.2$cm, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
Activation maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

@@

\paragraph{Linear autoencoder.}
The affine autoencoder takes the input 
$x$ and projects it into coordinates of a latent representation $z$
by

\begin{eqnarray}
  \begin{split}
  z &= W_0 x + b_{W_0} \\
  x' &= W_1 z + b_{W_1}
  \end{split}
  \label{eq:autoenc}
\end{eqnarray}

where $x \in \mathbb{R}^{d}$ denotes the vector of $d=79,941$
voxel values from each
rest map, $z \in \mathbb{R}^{n}$ is the $n$ hidden dimensions (i.e.,
distributed neural activity patterns), and 
$x' \in \mathbb{R}^{d}$ is the reconstruction vector of the original activity map
from the hidden variables. 
% XXX use bold letters for vectors and matrices
Further, ${W_0}$ denotes the weight matrix that
transforms
from input space into the hidden space (encoder),
${W_1}$ is the weight matrix for back-projection from the hidden variables to the
output space (decoder).
${b_{W_0}}$ and ${b_{W_1}}$ are bias vectors.
Note that ${W_0}$ and ${W_1}$ are tied,
$W_0$ = $W_1^T$.
The optimal model parameters ${W_0}, {b_{W_0}}, {b_{W_1}}$ are found by
minimizing the imperfection of reconstruction according to squared error

\begin{eqnarray}
  {\mathcal{L_{AE}}}(x, x') = || x - x' ||^2
\end{eqnarray}

@@

\paragraph{Reduced-rank logistic regression.}
Lossy compression by a low-dimensional bottleneck
is also imposed by the first layer of the low-rank
multinomial logistic regression.
It gives the probability of an input $x$ to belong
to a class $i \in \mathbb{N}$

\begin{eqnarray}
  \begin{split}
    P(Y=i|x, V_0,V_1,b_{V_0}, b_{V_1}) &= softmax_i(V_1 (V_0 x + b_{V_0}) + b_{V_1}) \\
    &= \frac {e^{V_{1i} (V_{0i} x + b_{V_{0i}}) + b_{V_{1i}}}} {\sum_j e^{V_{1j} (V_{0j} x + b_{V_{0j}}) + b_{V_{1j}}}}            
  \end{split}
  \label{eq:lr}
\end{eqnarray}

where the matrix $V_0 \in \mathbb{R}^{dxn}$ transforms the input $x \in \mathbb{R}^{d}$
into $n$ latent components
and the matrix $V_1 \in \mathbb{R}^{nxc}$ projects the latent components
onto hyperplanes that reflect $c$ class probabilities.
$b_{V_0}$ and $b_{V_1}$ are corresponding
bias vectors.
The loss function is given by

\begin{eqnarray}
  \begin{split}
    {\mathcal{L_{LR}}}(x, y) = \frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|x^{(i)}; V_1 (V_0 x + b_{V_0}) + b_{V_1})
\end{split}
\label{eq:lr_loss}
\end{eqnarray}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the low-rank logistic regression
are linked one two levels. First, their transformation matrices mapping from
input to the latent space are identical
\begin{eqnarray}
  V_0 = W_0
\end{eqnarray}
We thus search for a compression of the 79,941 voxel values into $n$ latent
components that represent an optimal latent code for both
rest and task activity data.
Second, the objectives of the autoencoder and the low-rank
logistic regression are interpolated in the common loss function

@@

In so doing, we search for the combined model parameters
$\theta=\{V_0,V_1,b_{V_0}, b_{V_1}, b_{W_0}, b_{W_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
For parameter shrinkage, we
impose a combination of $\ell_1$ and $\ell_2$ penalties
$\forall p \in \theta$ (i.e., elasticnet).

@@

\begin{eqnarray}
  \begin{split}
v^{(t+1)} &= \rho v^{(t+1)} + (1 - \rho)\left(\frac{\partial f}{\partial \theta}\right)^2
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla f(\theta^{(t)})}{\sqrt{v^{(t+1)} + \epsilon}},
  \end{split}
\end{eqnarray}

@@

\bigskip
% XXX the accuracy figure is not very nice. export is in pdf.
% XXX change the titles. remove the resolution information from the title.
% XXX larger legends. they're not readable currently
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.png}
\end{centering}
\caption{\textbf{Generalization performance}
Depicts in-sample (apply current model on training set, \textit{left column}),
out-of-sample (current model on test set, \textit{middle column}),
and out-of-dataset (compress other dataset using the
learned decomposition for classification, \textit{right column}) accuracy.
Curves for different choices of $\lambda$ as well as for ordinary
logistic regression (\textit{yellowish})
across iterations over the HCP task data (epochs).
XXX: Message ??
}
\end{figure}

@@

\begin{figure}
% XXX: figure not readable
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/support_bars.png}
\end{centering}
\caption{\textbf{Relation between support recovery and out-of-dataset performance}
\textit{Upper row:} the fit-for-purpose of the classification models
is quantified by linear correlation between the
z-scored average task image and the z-scored weights from
ordinary (\textit{red}) and present (\textit{blue colors}) logistic
regression ($\lambda=0.25, 0.50, 0.75, 1.00$).
\textit{Lower row:} ordinary logistic regression is applied to the
ARCHI dataset after projection into the latent dimensions learned from
the HCP task data.
}
\end{figure}

@@

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{figures/figure3.png}
  \end{center}
  \caption {\textbf{
  Equilibrium between autoencoder and low-rank logistic regression}
  One learned decomposition component (out of 20) between the only-autoencoder
  (\textit{uppermost panel}) and only-logistic-regression
  (\textit{lowermost panel}) scenarios.
  The congruent structure (\textit{red, middle column})
  in the
  posterior cingulate cortex, posterior mid-cingulate cortex, and medial
  prefrontal cortex appears to be important in decompositions of the rest and
  task neuroimaging data. With increasing weight on the supervised learning,
  non-congruent structure emerges in the early (\textit{blue}) and
  lateral (\textit{red}) visual cortex
  (\textit{right column}). Matrix weights were z-scored.
% XXX To me there is no clear message here. You can always interpolate between arbitrary maps... What does this bring to you ? 
 }
\end{figure}

\section{Discussion and Conclusion}
%
There is an increasing occurrence of high-dimensional problems in the
neuroimaging domain. 
%
% XXX: meaningless and useless for your work
%
Yet, there is also considerable uncertainty about the
most pertinent representation
of neural activation information.
This calls for new statistical learning algorithms that
behave well in large samples. Ideally, they should acknowledge
and exploit widely-accepted neuroscientific knowledge.
We here propose such an estimator that learns
dimensionality reduction
in a neurobiological and interpretable fashion.
% You really should emphasize the problem of accumulating data and knowledge which is much more interesting. Cite Yannick.

Using the flexibility of autoencoders, we
learn the optimal decomposition from high-dimensional
voxel brain space into the most important activation patterns for a
statistical learning question of interest.
The higher generalization accuracy and support recovery, comparing to
state-of-the-art classification, hold potential
for adoption in various every-day analyses in the neuroimaging field.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
This domain-specifc classification algorithms
encourages departure from the artificial and statistically
less attractive voxel space.
From a neurobiological perspective,
we demonstrate that neural activity underlying defined mental operations
can readily be explained by linear combinations of the main activation
patterns. That is,
neural activity data as measured by fMRI probably concentrate near
a low-dimensional manifold of brain networks.
These fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
Moreover, data-intense neuroimaging studies in the future can be made tractable
by reduction to the neurobiological essence.
This can be particulary useful in multi-center clinical studies
on the prediction of disease trajectories and drug treatment responses
in psychiatric and neurological populations.
We hope that these first steps stimulate development towards
powerful semi-supervised classification methods.

\paragraph{Acknowledgment}
{\small
Data were provided the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.)}

>>>>

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.
%
Existing neuroimaging methods typically perform either discovery of unknown
neural structure or testing of neural structure associated with mental tasks.
%
However, testing hypotheses on the neural correlates of mental tasks
depends on the representation of the observations.
%
We therefore propose to optimize
representation and task classification in
an identical statistical learning problem.
%
We show that this approach yields more accurate and more interpretable
neural models of psychological tasks in a reference dataset.
%

@@

\section{Introduction}
%
Methods for neuroimaging research can be grouped by discovering
neurobiological structure or assessing the neural correlates associated
with mental tasks.
To discover, on the one hand, spatial distributions of neural activity
structure across time,
independent component analysis (ICA) \cite{beckmann2005} is often used.
It decomposes the BOLD (blood-oxygen level-dependent) signals into the
primary modes of variation.
The ensuing spatial activity patterns are believed to represent
brain networks of
functionally interacting regions \cite{smith2009}.
Similarly, sparse principal component analysis (SPCA) \cite{varoqu2011}
has been used to
separate BOLD signals into parsimonious network components.
Thus extracted brain networks are probably
manifestations of electrophysiological oscillation frequencies \cite{hipp15}.
Their fundamental organizational role is further
attested by continued covariation during sleep and anesthesia \cite{fox07}.
%
Network discovery by applying ICA or SPCA is typically performed on
task-unrelated (i.e., \textit{unlabeled}) ``resting-state'' data.
These capture brain dynamics
during ongoing random thought without controlled environmental stimulation.
In fact, a large proportion of the BOLD signal is known
not to correlate with a particular behavior, stimulus, or experimental task
\cite{fox07}.

To test, on the other hand,
the neural correlates underlying mental tasks,
the general linear model (GLM) is the dominant approach \cite{friston94}.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI)
elucidate the influence of one brain region on another conditioned
by experimental tasks \cite{friston97}.
As a last example, always more neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to task-associated (i.e., \textit{labeled})
data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses typically yield
results in a voxel space.
This ignores the fact that the BOLD
signal exhibits spatially distributed activity patterns.
%
Second, existing supervised neuroimaging analyses
do not acknowledge the abundance
of easily acquired resting-state data \cite{biswaldiscovery}.
These may allow better sampling the manifold of brain states
due to the high task-rest similarities of neural activity patterns,
as observed using ICA \cite{smith2009}.

Both these biological properties can be conjointly exploited in a mixed
 (i.e., use task and rest data)
low-rank (i.e., perform network decomposition)
approach.
%
The integration of brain-network discovery into
supervised classification formally yields a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
Autoencoders
are one-layered learning models that condense the input data to
local and global representations
by improving reconstruction from them.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
%
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \caption {\textbf{Architecture. XXX develop the cpation. Main msg ?}
  }
\end{wrapfigure}
%
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex transformation \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Hence, autoencoders can flexibly project the neuroimaging data
onto the main axes of variation and thus
reverse-engineer properties of the underlying
neural processes \cite{olshausen96}.

In the present investigation,
a linear autoencoder will be fed by (unlabeled) rest data and
integrated as bottleneck
into a low-rank logistic regression fed by (labeled) task data.
We can then solve the unsupervised data representation and the
supervised classification in an identical statistical learning
problem.
%
From the perspective of dictionary learning, the first layer represents
the discovered set of basis functions
whose linear combinations are learned
by the second layer \cite{olshausen96}.
%
Neurobiologically, this allows 
delineating a low-dimensional manifold of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
put probability mass on the most neurobiologically
valid models in the hypothesis set.
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.

\section{Methods}
%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose the Human Connectome Project (HCP) resources
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500 unrelated,
healthy HCP participants.
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants.
The task paradigms include
1) working memory/cognitive control processing, 2)
incentive processing, 3) visual and somatosensory-motor processing,
4) language processing (semantic and phonological processing),
5) social cognition, 6) relational processing, and 7) emotional
processing. All data were acquired on the same Siemens Skyra 3T scanner.
Whole-brain EPI acquisitions were acquired with a
32 channel head coil (TR=720ms, TE=33.1ms, flip angle=52°, BW=2290Hz/Px,
in-plane FOV=$280\times180$mm, 72 slices, 2.0mm isotropic voxels).
The ``minimally preprocessed'' pipeline includes
gradient unwarping, motion correction, fieldmap-based EPI distortion
correction, brain-boundary-based registration of EPI to structural
T1-weighted scans, nonlinear (FNIRT) registration into MNI space,
and grand-mean intensity normalization. Activity maps were spatially
smoothed with a Gaussian kernel of 4mm (FWHM). A GLM was
implemented by FILM from the FSL suite with model regressors from convolution
with a “canonical” hemodynamic response function and from temporal derivatives.
HCP tasks were conceived to modulate activity
in a maximum of different brain regions and neural systems. Indeed, at
least 70\% of the participants showed consistent brain activity in
contrasts from the task battery, which certifies excellent
coverage \cite{barch2013}.
%
% No, this is not what is called 'coverage'
%
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common 60x72x60 space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supservised analyses were based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

@@

We were further interested in the utility of the optimal low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classifcation problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$19.2\times19.2$cm, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
activity maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

@@

\paragraph{Linear autoencoder.}
The affine autoencoder takes the input 
$\mathbf{x}$ and projects it into coordinates of a
latent representation $\mathbf{z}$
by

\begin{eqnarray}
  \begin{split}
    \mathbf{z} &= \mathbf{W_0} \mathbf{x} + \mathbf{b_{W_0}} \\
    \mathbf{x'} &= \mathbf{W_1} \mathbf{z} + \mathbf{b_{W_1}}
  \end{split}
  \label{eq:autoenc}
\end{eqnarray}

where $\mathbf{x \in \mathbb{R}^{d}}$ denotes the vector of $d=79,941$
voxel values from each
rest map,
$\mathbf{z \in \mathbb{R}^{n}}$ is the $n$ hidden dimensions (i.e.,
distributed neural activity patterns), and 
$\mathbf{x' \in \mathbb{R}^{d}}$
is the reconstruction vector of the original activity map
from the hidden variables. 
% XXX use bold letters for vectors and matrices
Further, $\mathbf{W_0}$ denotes the weight matrix that
transforms
from input space into the hidden space (encoder),
$\mathbf{W_1}$ is the weight matrix for back-projection
from the hidden variables to the
output space (decoder).
$\mathbf{b_{W_0}}$ and $\mathbf{b_{W_1}}$ are bias vectors.
Note that $\mathbf{W_0}$ and $\mathbf{W_1}$ are tied,
$\mathbf{W_0} = \mathbf{W_1^T}$.
The optimal model parameters $\mathbf{W_0, b_{W_0}, b_{W_1}}$ are found by
minimizing the imperfection of reconstruction according to squared error

\begin{eqnarray}
  {\mathcal{L_{AE}}}(\mathbf{x}, \mathbf{x'}) = || \mathbf{x} - \mathbf{x'} ||^2
\end{eqnarray}

@@

\paragraph{Reduced-rank logistic regression.}
Lossy compression by a low-dimensional bottleneck
is also imposed by the first layer of the low-rank
multinomial logistic regression.
It gives the probability of an input $\mathbf{x}$ to belong
to a class $i \in \mathbb{N}$

\begin{eqnarray}
  \begin{split}
    P(Y=i|\mathbf{x, V_0,V_1,b_{V_0}, b_{V_1}}) &= softmax_i(\mathbf{V_1 (V_0 x + b_{V_0}) + b_{V_1}}) \\
    &= \frac {e^{\mathbf{V_{1i} (V_{0i} x + b_{V_{0i}}) + b_{V_{1i}}}}} {\sum_j e^{\mathbf{V_{1j} (V_{0j} x + b_{V_{0j}}) + b_{V_{1j}}}}}            
  \end{split}
  \label{eq:lr}
\end{eqnarray}

where the matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxc}}$
projects the latent components
onto hyperplanes that reflect $c$ class probabilities.
$\mathbf{b_{V_0}}$ and $\mathbf{b_{V_1}}$ are corresponding
bias vectors.
The loss function is given by

\begin{eqnarray}
  \begin{split}
    {\mathcal{L_{LR}}}(\mathbf{x, y}) = \frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|\mathbf{x^{(i)}; V_1 (V_0 x + b_{V_0}) + b_{V_1}})
\end{split}
\label{eq:lr_loss}
\end{eqnarray}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the low-rank logistic regression
are linked one two levels. First, their transformation matrices mapping from
input to the latent space are identical
\begin{eqnarray}
  \mathbf{V_0} = \mathbf{W_0}
\end{eqnarray}
We thus search for a compression of the 79,941 voxel values into $n$ latent
components that represent an optimal latent code for both
rest and task activity data.
Second, the objectives of the autoencoder and the low-rank
logistic regression are interpolated in the common loss function

@@

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,b_{V_0}, b_{V_1}, b_{W_0}, b_{W_1}}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
For parameter shrinkage, we
impose a combination of $\ell_1$ and $\ell_2$ penalties
$\forall p \in \theta$ (i.e., elasticnet).

@@

\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t+1)}} + (1 - \rho)\left(\frac{\partial f}{\partial \theta}\right)^2
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla f(\theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}

@@

\bigskip
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.png}
\end{centering}
\caption{\textbf{Generalization performance}
Depicts in-sample (apply current model on training set, \textit{left column}),
out-of-sample (current model on test set, \textit{middle column}),
and out-of-dataset (compress other dataset using the
learned decomposition for classification, \textit{right column}) accuracy.
Curves for different choices of $\lambda$ as well as for ordinary
logistic regression (\textit{yellowish})
across iterations over the HCP task data (epochs).
XXX: Message ??
}
\end{figure}

@@

\begin{figure}
% XXX: figure not readable
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/semisup.png}
\end{centering}
\caption{\textbf{Relation between support recovery and out-of-dataset performance}
\textit{Upper row:} the fit-for-purpose of the classification models
is quantified by linear correlation between the
z-scored average task image and the z-scored weights from
ordinary (\textit{red}) and present (\textit{blue colors}) logistic
regression ($\lambda=0.25, 0.50, 0.75, 1.00$).
\textit{Lower row:} ordinary logistic regression is applied to the
ARCHI dataset after projection into the latent dimensions learned from
the HCP task data.

\textbf{whenever I tried to have 3+ times more rest than task images results dropped dramatically...naive interpretation: if the rest information is too rich it overrules task patterns and hinders the supervised learning
}
}
\end{figure}

@@

the features of the classes show that we captured activaiton properties,
not particpant or paradigm properties

@@

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of autoencoders, we
learn the optimal decomposition from high-dimensional
voxel brain space into the most important activity patterns for a
supervised learning question.
The higher generalization accuracy and support recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
This domain-specifc classification algorithms
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
neural activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns. That is,
fMRI data probably concentrate near
a low-dimensional manifold of brain networks.
These fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
We hope that these first steps stimulate development towards
powerful semi-supervised classification methods in systems neuroscience.

% grand perspective
In the future, automatic reduction of brain images to
their neurobiological essence
may facilitate data-intense neuroimaging studies.
Initiatives for data collection are rapidly increasing
in neuroscience \cite{poldrack2014data}.
These promise integration
of accumulating neuroscientific knowledge from neuroimaging databases.
Tractability by
condensed feature representations can avoid the ill-posed problem of
learning the full distribution of activity patterns.
%
This is relevant to the 
multi-class challenges spanning the human cognitive space
\cite{schwartz2013mapping}
and
combination of neuroimaging with
high-throughput genetic analyses, such as GWAS
\cite{need2010gwas}.
%
The biggest socioeconomic potential may
lie in across-hospital clinical studies that
predict disease trajectories and drug responses
in psychiatric/neurological populations
\cite{frackowiak2015future, gustav2011cost}.

\paragraph{Acknowledgment}
{\small
Data were provided the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.).
}
